---
title: "RCV Package Workflow"
author: "Jay Lee, Matthew Yancheff"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(rcv)
library(dplyr)
```

`rcv` helps you work directly with raw ballot image and cast vote record data to run elections.

__Features__

* Read in ballot image and master lookup files
* Merge these files to get a "readable" ballot layout
* Conduct elections, and view a round-by-round table of results
* Visualize the flow of voters with an interactive Sankey diagram
* Compatible with `dplyr`/`magrittr` pipe syntax (`%>%`)

You can install the development version of `rcv` here:

```{r eval = F}
devtools::install_github("ds-elections/rcv")
```

Created by:

- Jay Lee, jaylee@reed.edu
- Matthew Yancheff, yanchefm@reed.edu

The style of this vignette is inspired by the [googlesheets]("github.com/jennybc/googlesheets") R package.

## Basic Workflow Demo

`sf_bos_ballot` is included as an example raw ballot image, and `sf_bos_lookup` as an example raw master lookup. Both are included as `.rda`s, and they are in the "WinEDS" format. This data comes from the 2016 San Francisco Board of Supervisors elections ([San Francisco Department of Elections](http://www.sfelections.org/results/20161108/#english_detail)).

```{r}
head(sf_bos_ballot)
head(sf_bos_lookup)
```

#### Cleaning Data

The streamlined version of this process is done with the `clean_ballot()` function. `b_header` and `l_header` are logical values, based on whether the ballot and lookup file respectively have a header for the first row.

```{r}
cleaned <- clean_ballot(ballot = sf_bos_ballot, b_header = T, 
                        lookup = sf_bos_lookup, l_header = T, 
                        format = "WinEDS")
knitr::kable(head(readable(cleaned)))
```

To access intermediate steps, the following process can be used.

```{r}
# Import and label ballot image
a <- import_data(data = sf_bos_ballot, header = T) %>%
    label(image = "ballot", format = "WinEDS")

# Import and label master lookup
b <- import_data(data = sf_bos_lookup, header = T) %>%
    label(image = "lookup", format = "WinEDS")

# Merge these two tables
c <- characterize(ballot = a, lookup = b, format = "WinEDS")

knitr::kable(head(readable(c)))
```

The `readable()` function takes the clean image, which is formatted for ease in computation, and formats it to be easily read manually.

#### Running Elections

This is done with the `rcv_tally()` function. `sf_bos_clean` is included as an example of a pre-cleaned ballot using the functions above. We will run the District 1 election from this ballot image.

```{r}
results <- rcv_tally(sf_bos_clean, "Board of Supervisors, District 1")
knitr::kable(results)
```

Sandra Lee Fewer wins in Round 9, with 14,705 votes to Marjan Philhour's 13,126. 4,360 votes were left blank, marked invalid, or exhausted in this election.

#### Visualizing Data

We have two recommended methods of visualizing RCV data. Both utilize a flowchart called a "Sankey diagram" to show the transfer of voters between rounds. We will use each method to visualize the transfer of voters in the San Francisco District 7 Board of Supervisors election, because District 1 has too many crossings to be readable.

Method 1 (preferred because it is interactive, quicker, and more readable) uses the `networkD3` package:

```{r}
d3_7 <- rcv::make_d3list(results = sf_7_results)
networkD3::sankeyNetwork(Links = d3_7$values, Nodes = d3_7$names,
                         Source = "source", Target = "target",
                         Value = "value", NodeID = "candidate", units = "voters",
                         fontSize = 12, nodeWidth = 20, width = 600)

```


Method 2 uses the `alluvial` package (this type of graphic is also called an alluvial diagram):

```{r}
alluvial_7 <- rcv::make_alluvialdf(image = sf_bos_clean,
                              rcvcontest = "Board of Supervisors, District 7")
alluvial::alluvial(
  alluvial_7[,1:4], 
  freq = alluvial_7$frequency,
  col = ifelse(alluvial_7$round4 == "NORMAN YEE", "lightgreen", "gray"),
  border = "gray",
  alpha = 0.7,
  blocks = TRUE
)
```

